### 2018-05-08 submission/manuscript_R1.Rmd
final changes

An increasing body of literature suggests that both individual and collections of bacteria are associated with the progression of colorectal cancer. As the number of studies investigating these associations increases and the number of subjects in each study increases, a meta-analysis to identify the associations that are the most predictive of disease progression is warranted. We analyzed previously published 16S rRNA gene sequencing data collected from feces and colon tissue. We quantified the odds ratios (ORs) for individual bacterial taxa that were associated with an individual having tumors relative to a normal colon. Among the fecal samples, there were no taxa that had significant ORs associated with adenoma and there were `r length(crc_OR_stool$measure)` taxa with significant ORs associated with carcinoma. Similarly, among the tissue samples, there were no taxa that had a significant OR associated with adenoma and there were `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs associated with carcinoma. Among the significant ORs, the association between individual taxa and tumor diagnosis was equal or below `r ifelse(max(crc_OR_stool$rr) > max(crc_unmatched_OR_tissue$rr), round(max(crc_OR_stool$rr), digits = 2), round(max(crc_unmatched_OR_tissue$rr), digits = 2))`. Because individual taxa had limited association with tumor diagnosis, we trained Random Forest classification models using only the taxa that had significant ORs, using the entire collection of taxa found in each study, and using operational taxonomic units defined based on a 97% similarity threshold. All training approaches yielded similar classification success as measured using the Area Under the Curve. The ability to correctly classify individuals with adenomas was poor and the ability to classify individuals with carcinomas was considerably better using sequences from fecal or tissue.

@@

The authors would like to thank all the study participants who were a part of each of the individual studies analyzed. We would also like to thank each of the study authors for making their sequencing reads and metadata available for use. Finally, we would like to thank the members of the Schloss lab for their valuable feedback and proofreading during the formulation of this manuscript.

>>>>

An increasing body of literature suggests that both individual and collections of bacteria are associated with the progression of colorectal cancer. As the number of studies investigating these associations increases and the number of subjects in each study increases, a meta-analysis to identify the associations that are the most predictive of disease progression is warranted. We analyzed previously published 16S rRNA gene sequencing data collected from feces and colon tissue. We quantified the odds ratios (ORs) for individual bacterial taxa that were associated with an individual having tumors relative to a normal colon. Among the fecal samples, there were no taxa that had significant ORs associated with adenoma and there were `r length(crc_OR_stool$measure)` taxa with significant ORs associated with carcinoma. Similarly, among the tissue samples, there were no taxa that had a significant OR associated with adenoma and there were `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs associated with carcinoma. Among the significant ORs, the association between individual taxa and tumor diagnosis was equal or below `r ifelse(max(crc_OR_stool$rr) > max(crc_unmatched_OR_tissue$rr), round(max(crc_OR_stool$rr), digits = 2), round(max(crc_unmatched_OR_tissue$rr), digits = 2))`. Because individual taxa had limited association with tumor diagnosis, we trained Random Forest classification models using only the taxa that had significant ORs, using the entire collection of taxa found in each study, and using operational taxonomic units defined based on a 97% similarity threshold. All training approaches yielded similar classification success as measured using the Area Under the Curve. The ability to correctly classify individuals with adenomas was poor and the ability to classify individuals with carcinomas was considerably better using sequences from feces or tissue.

@@

The authors would like to thank all the study participants who were a part of each of the individual studies analyzed. We would also like to thank each of the study authors for making their sequencing reads and metadata available for use. Finally, we would like to thank the members of the Schloss lab for their valuable feedback and proofreading during the formulation of this manuscript. Salary support for Marc A. Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants P30DK034933 and 1R01CA215574.

### 2018-05-08 submission/manuscript_R1.tex
final changes

An increasing body of literature suggests that both individual and
collections of bacteria are associated with the progression of
colorectal cancer. As the number of studies investigating these
associations increases and the number of subjects in each study
increases, a meta-analysis to identify the associations that are the
most predictive of disease progression is warranted. We analyzed
previously published 16S rRNA gene sequencing data collected from feces
and colon tissue. We quantified the odds ratios (ORs) for individual
bacterial taxa that were associated with an individual having tumors
relative to a normal colon. Among the fecal samples, there were no taxa
that had significant ORs associated with adenoma and there were 8 taxa
with significant ORs associated with carcinoma. Similarly, among the
tissue samples, there were no taxa that had a significant OR associated
with adenoma and there were 3 taxa with significant ORs associated with
carcinoma. Among the significant ORs, the association between individual
taxa and tumor diagnosis was equal or below 7.11. Because individual
taxa had limited association with tumor diagnosis, we trained Random
Forest classification models using only the taxa that had significant
ORs, using the entire collection of taxa found in each study, and using
operational taxonomic units defined based on a 97\% similarity
threshold. All training approaches yielded similar classification
success as measured using the Area Under the Curve. The ability to
correctly classify individuals with adenomas was poor and the ability to
classify individuals with carcinomas was considerably better using
sequences from fecal or tissue.

@@

\textbf{\emph{Statistical Analysis.}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package (54). For
OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package (55) and
Z-score normalized using the car (v2.1.6) package (56). Testing for OTU
richness, evenness, and Shannon diversity differences utilized linear
mixed-effect models to correct for study, repeat sampling of individuals
(tissue only), and 16S rRNA gene sequence region used using the lme4
(v1.1.15) package (57). ORs were analyzed using both the epiR (v0.9.93)
and metafor (v2.0.0) packages (58, 59) by assessing how many individuals
with and without disease were above and below the overall median value
within each specific study. OR significance testing utilized the
chi-squared test. Community structure differences were calculated using
the Bray-Curtis dissimilarity index and PERMANOVA was used to test for
tumor-associated differences in structure with the vegan (v2.4.5)
package (60). Random Forest models were built using both the caret
(v6.0.78) and randomForest (v4.6.12) packages (61, 62). All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages (63,
64).

>>>>

An increasing body of literature suggests that both individual and
collections of bacteria are associated with the progression of
colorectal cancer. As the number of studies investigating these
associations increases and the number of subjects in each study
increases, a meta-analysis to identify the associations that are the
most predictive of disease progression is warranted. We analyzed
previously published 16S rRNA gene sequencing data collected from feces
and colon tissue. We quantified the odds ratios (ORs) for individual
bacterial taxa that were associated with an individual having tumors
relative to a normal colon. Among the fecal samples, there were no taxa
that had significant ORs associated with adenoma and there were 8 taxa
with significant ORs associated with carcinoma. Similarly, among the
tissue samples, there were no taxa that had a significant OR associated
with adenoma and there were 3 taxa with significant ORs associated with
carcinoma. Among the significant ORs, the association between individual
taxa and tumor diagnosis was equal or below 7.11. Because individual
taxa had limited association with tumor diagnosis, we trained Random
Forest classification models using only the taxa that had significant
ORs, using the entire collection of taxa found in each study, and using
operational taxonomic units defined based on a 97\% similarity
threshold. All training approaches yielded similar classification
success as measured using the Area Under the Curve. The ability to
correctly classify individuals with adenomas was poor and the ability to
classify individuals with carcinomas was considerably better using
sequences from feces or tissue.

@@

\textbf{\emph{Statistical Analysis.}} All statistical analysis after
sequence processing utilized the R (v3.4.4) software package (54). For
OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package (55) and
Z-score normalized using the car (v2.1.6) package (56). Testing for OTU
richness, evenness, and Shannon diversity differences utilized linear
mixed-effect models to correct for study, repeat sampling of individuals
(tissue only), and 16S rRNA gene sequence region used using the lme4
(v1.1.15) package (57). ORs were analyzed using both the epiR (v0.9.93)
and metafor (v2.0.0) packages (58, 59) by assessing how many individuals
with and without disease were above and below the overall median value
within each specific study. OR significance testing utilized the
chi-squared test. Community structure differences were calculated using
the Bray-Curtis dissimilarity index and PERMANOVA was used to test for
tumor-associated differences in structure with the vegan (v2.4.5)
package (60). Random Forest models were built using both the caret
(v6.0.78) and randomForest (v4.6.12) packages (61, 62). All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages (63,
64).
